---
title: "lab 9 - Monte Carlo Methods"
subtitle: "BSMM 8740 Fall 2024"
author: "Thi Bao Han Truong"
format: html
editor: visual
self-contained: true
---

## Introduction

In today's lab, you'll practice sampling from distributions and working with Markov chains.

## Getting started

-   To complete the lab, log on to **your** github account and then go to the class [GitHub organization](https://github.com/bsmm-8740-fall-2024) and find the **2024-lab-9-\[your github username\]** repository .

    Create an R project using your **2024-lab-9-\[your github username\]** repository (remember to create a PAT, etc.) and add your answers by editing the `2024-lab-9.qmd` file in your repository.

-   When you are done, be sure to: **save** your document, **stage**, **commit** and [**push**]{.underline} your work.

::: callout-important
To access Github from the lab, you will need to make sure you are logged in as follows:

-   username: **.\\daladmin**
-   password: **Business507!**

Remember to (create a PAT and set your git credentials)

-   create your PAT using `usethis::create_github_token()` ,
-   store your PAT with `gitcreds::gitcreds_set()` ,
-   set your username and email with
    -   `usethis::use_git_config( user.name = ___, user.email = ___)`
:::

## Packages

```{r}
#| echo: false
#| message: false
#| warning: false
# check if 'librarian' is installed and if not, install it
if (! "librarian" %in% rownames(installed.packages()) ){
  install.packages("librarian")
}
  
# load packages if not already loaded
librarian::shelf(expm, ggplot2)
theme_set(theme_bw(base_size = 18) + theme(legend.position = "top"))
```

## Exercise 1: Markov Chains

Here is a four-state Markov chain that could model customer loyalty for a subscription-based service, with one month between steps in the chain.

[**States**]{.underline}:

-   State A (New Customer): The customer has just signed up.
-   State B (Engaged Customer): The customer is actively using the service and seems satisfied.
-   State C (At-Risk Customer): The customer is showing signs of disengagement (e.g., reduced usage or negative feedback).
-   State D (Churned Customer): The customer has canceled their subscription.

[**Transition Probabilities**]{.underline}:

-   From State A (New Customer), there’s a high chance the customer either becomes engaged (State B) or starts showing signs of disengagement (State C).
-   From State B (Engaged Customer), there’s a probability of remaining engaged or transitioning to at-risk (State C), and a smaller probability of churning (State D).
-   From State C (At-Risk Customer), the customer may either re-engage (return to State B) or churn (State D).
-   From State D (Churned Customer), it's possible the company might re-acquire the customer through marketing efforts, which would move them back to State A.

This type of Markov model can help businesses predict customer behavior, optimize marketing efforts, and focus on retention strategies.

What is the probability that a customer that has [just signed up]{.underline} is [still a customer after 6 months]{.underline}?

::: {.callout-note appearance="simple" icon="false"}
## YOUR ANSWER:

```{r}
#| echo: true
#| eval: false
# the transition matrix is
P <- 
  matrix(
    c(0, 0.6, 0.4, 0,
      0, 0.75, 0.25, 0,
      0, 0.5, 0, 0.5,
      0.3, 0, 0, 0.7
      )
    , nrow =4, byrow = TRUE
  )

# use %^% from the expm package to compute the k-th power of a matrix (k = 6 months)
initial_state <- c(1, 0, 0, 0)
state_after_6_months <- initial_state %*% (P %^% 6)

# sum the probabilities of the non-churned customer states after 6 steps
prob_still_customer <- sum(state_after_6_months[1:3])

prob_still_customer
```

The probability that a customer that has just signed up is still a customer after 6 months is 74.%
:::

## Exercise 2: Markov Chains

A simpler customer churn model for each monthly period is as follows:

-   a current subscriber cancels their subscription with probability 0.2
-   a current non-subscriber starts their subscription with probability with probability 0.06

write the state transition matrix $\mathsf{P}_{i,j}$, and compute the stationary distribution $\pi$ for this Markov Chain, confirming that $\pi\mathsf{P}=\pi$ and that the sum of the elements of $\pi$ equals $1.0$.

What percent of customers remain once the chain has reached the steady state?

::: {.callout-note appearance="simple" icon="false"}
## YOUR ANSWER:

```{r}
#| echo: true
#| eval: false
# replace the placeholders '_' with the state transition probabilities
P <- 
  matrix(
    c(0.8, 0.2,
     0.06, 0.94
      )
    , nrow =2, byrow = TRUE
  )
```

```{r}
#| echo: true
#| eval: false
# compute transpose(I-P) and add a row of 1's to the bottom 
# call the resulting matrix A
I <- diag(2)
A <- rbind(t(I - P), c(1, 1))

# create a vector called b with the # of elements equal to the number of rows of A
# with elements all zero but the last one
b <- c(0, 0, 1)
  
# compute pi by solving (A x pi) = b using qr.solve
pi <- qr.solve(A, b)

# confirm (pi x P) = pi and 
pi_P <- pi %*% P
identical(round(pi, 10), round(pi_P, 10))
  
# confirm pi[1] + pi[2] == 1
sum_pi <- sum(pi)
identical(round(sum_pi, 10), 1.0)

percent_remaining <- pi[1] * 100
percent_remaining
```

In the steady state, the probability of being a current customer is 23.07%
:::

## Exercise 3: Acceptance probability

We want to sample from the Poisson distribution $\mathbb{P}(X=x)\sim \lambda^xe^{-\lambda}/x!$ using a Metropolis Hastings algorithm.

For the proposal we toss a fair coin and add or subtract 1 from $x$ to obtain $y$ as follows:

$$
q(y|x)=\begin{cases}
\frac{1}{2} & x\ge1,\,y=x\pm1\\
1 & x=0,\,y=1\\
0 & \mathrm{otherwise}
\end{cases}
$$ show that the acceptance probability is

$$
\alpha(y|x)=\begin{cases}
\min\left(1,\frac{\lambda}{x+1}\right) & x\ge1,\,y=x+1\\
\min\left(1,\frac{x}{\lambda}\right) & x\ge2,\,y=x-1
\end{cases}
$$

and $\alpha(1|0)=\min(1,\lambda/2)$, $\alpha(0|1)=\min(1,2/\lambda)$

::: {.callout-note appearance="simple" icon="false"}
## YOUR ANSWER:
:::

## Exercise 4: Samples from Poisson pmf

Given the following function for the acceptance probability

```{r}
alpha <- function(y,x, lambda){
  if(x >= 1 & y == x+1){
    min(1,lambda/(x+1))
  }else if(x >= 2 & y == x-1){
    min(1,x/lambda)
  }else if(x == 0 & y == 1){
   min(1,lambda/2)
  }else{
    min(1,2/lambda)
  }
}
```

1.  Write a MH algorithm to draw 2000 samples from a from a Poisson pmf with $\lambda = 20$ starting from $x_0=1$.

2.  Compare the sample quantiles at probabilities c(0.1,.25,0.5, 0.75, 0.9) with the theoretical quantiles for the Poisson distribution (using the `qpois` function)

::: {.callout-note appearance="simple" icon="false"}
## YOUR ANSWER:

```{r}
# MH algorithm drawing samples from a Poisson(20) pmf
MH_sampler <- function(lambda, n_samples, x0) {
  samples <- numeric(n_samples)
  x <- x0
  for (i in 1:n_samples) {
    # Proposal step: toss a fair coin to propose y = x + 1 or y = x - 1
    if (x == 0) {
      y <- 1
    } else {
      y <- ifelse(runif(1) < 0.5, x + 1, x - 1)
    }
    
    # Acceptance step
    if (runif(1) < alpha(y, x, lambda)) {
      x <- y
    }
    # Store the sample
    samples[i] <- x
  }
  return(samples)
}

# comparison of quantiles between the samples and the theoretical.
# Set parameters
lambda <- 20
n_samples <- 2000
x0 <- 1

# Generate samples
set.seed(123) # For reproducibility
samples <- MH_sampler(lambda, n_samples, x0)

# Compare sample quantiles with theoretical quantiles
probs <- c(0.1, 0.25, 0.5, 0.75, 0.9)
sample_quantiles <- quantile(samples, probs)
theoretical_quantiles <- qpois(probs, lambda)

# Display the comparison
comparison <- data.frame(
  Probability = probs,
  Sample_Quantiles = sample_quantiles,
  Theoretical_Quantiles = theoretical_quantiles
)

print(comparison)

```
:::

## Exercise 5: A Loan Portfolio

Our client is a bank with both asset and liability products in retail bank industry. Most of the bank's assets are loans, and these loans generate the majority of the total revenue earned by the bank. Hence, it is essential for the bank to understand the proportion of loans that have a high propensity to be paid in full and those which will finally become Bad loans.

All the loans that have been issued by the bank are classified into one of four categories/states :

1.  **Good Loans** : These are the loans which are in progress but are given to low risk customers. We expect most of these loans will be paid up in full with time.
2.  **Risky loans** : These are also the loans which are in progress but are given to medium or high risk customers. We expect a good number of these customers will default.
3.  **Bad loans** : The customer to whom these loans were given have already defaulted.
4.  **Paid up loans** : These loans have already been paid in full.

Your research has suggested the following state transition matrix for the bank loans

```{r}
# the 1-year state transition matrix for loans is:
P <- 
  matrix(
    c(0.7, 0.05, 0.03, 0.22,
      0.05, 0.55, 0.35, 0.05,
      0, 0, 1, 0, 
      0, 0, 0, 1
      )
    , nrow =4, byrow = TRUE
  )
```

Answer the following questions, given that the bank's records indicate 60% of the loans on the books are 'good loans' and 40% are 'risky loans'.

::: {.callout-note appearance="simple" icon="false"}
## YOUR ANSWER:

```{r}
# describe the current loan portfolio by state at the end of one year and two years.  
initial_state <- c(0.6, 0.4, 0, 0)
state_1_year <- initial_state %*% P

# describe the current loan portfolio by state at the end of two years.
state_2_years <- state_1_year %*% P
```

```{r}
# What percentage of good loans are paid in full after 20 years
# Compute the state distribution after 20 years for Good Loans and Risky Loans
initial_good_loans <- c(1, 0, 0, 0) 
initial_risky_loans <- c(0, 1, 0, 0)

P_20 <- P %^% 20
```

53.74 percent of good loans are paid in full after 20 years

```{r}
# What percentage of good loans are paid in full after 20 years
state_good_loans_20_years <- initial_good_loans %*% P_20
state_risky_loans_20_years <- initial_risky_loans %*% P_20

# Extract the percentages of Paid Loans
percentage_good_paid <- state_good_loans_20_years[4] * 100
percentage_risky_paid <- state_risky_loans_20_years[4] * 100


# Display the results
cat("Loan Portfolio After 1 Year:\n")
print(state_1_year)

cat("\nLoan Portfolio After 2 Years:\n")
print(state_2_years)

cat("\nPercentage of Good Loans Paid in Full After 20 Years:", percentage_good_paid, "%\n")
cat("Percentage of Risky Loans Paid in Full After 20 Years:", percentage_risky_paid, "%\n")

```

19.59 percent of risk loans are paid in full after 20 years
:::

::: render-commit-push
You're done and ready to submit your work! **Save**, **stage**, **commit**, and **push** all remaining changes. You can use the commit message "Done with Lab 6!" , and make sure you have committed and pushed all changed files to GitHub (your Git pane in RStudio should be empty) and that **all** documents are updated in your repo on GitHub.
:::

::: callout-important
## Submission

I will pull (copy) everyone's repository submissions at 5:00pm on the Sunday following class, and I will work only with these copies, so anything submitted after 5:00pm will not be graded. (**don't forget to commit and then push your work!**)
:::

## Grading

Total points available: 30 points.

| Component | Points |
|-----------|--------|
| Ex 1 - 5  | 30     |
